The curious world of constructive mathematics

### 2022-09-05 algar-lecture-notes.tex
Decidability of equality and minima of sets of natural numbers

\begin{ex}
Assuming that we cannot find our apartment key, but know that it has to be
somewhere (because we used it last night to unlock the door), constructively we
can only defend the doubly negated statement
\[ \neg\neg (\exists x\_ \text{the key is at position~$x$}). \]
\end{ex}

@@

@@

@@

@@

\begin{exercise}[Stability of the axioms]
Peano arithmetic (\textsc{pa}) is set in the language~$(0,S,+,\cdot)$ and has
the following axioms (where leading universal quantifiers are supressed for
brevity):
\begin{enumerate}
\renewcommand{\theenumi}{\arabic{enumi}}
\item $Sx \neq 0$
\item $Sx = Sy \Rightarrow x = y$
\item $y = 0 \vee (\exists x\_ y = Sx)$
\item $x + 0 = x$
\item $x + Sy = S(x+y)$
\item $x \cdot 0 = 0$
\item $x \cdot Sy = (x \cdot y) + x$
\item $P(0) \wedge (\forall n\_ P(n) \Rightarrow P(Sn)) \Longrightarrow
(\forall n\_ P(n))$ (one axiom for each formula~$P(n)$)
\end{enumerate}
Heyting arithmetic~(\textsc{ha}) has exactly the same axioms, but is based on intuitionistic
logic instead of classical logic.
\begin{enumerate}
\item Show that~\textsc{ha} proves the double-negation translation of each
axiom of~\textsc{pa}.
\item Convince yourself that~\textsc{izf} does not prove the double-negation
translation of the axiom of choice. Hence the double-negation translation
alone is insufficient to extract constructive content from proofs using the
axiom of choice.
\end{enumerate}
\end{exercise}

@@

\begin{exercise}[Minima of sets of natural numbers]
In Theorem~\ref{thm:min}(b) we showed that every inhabited set of natural
numbers does \notnot contain a minimal element. Explain how this result follows
from the double-negation embedding applied to the classical fact that every
inhabited set of natural numbers directly contains a minimal element.

{\noindent\scriptsize\emph{Hint.} It is not enough to just cite the
double-negation embedding. The resulting constructive theorem has to be
massaged a bit to yield Theorem~\ref{thm:min}(b).\par}
\end{exercise}

@@

>>>>

\begin{ex}\label{ex:key}
Assuming that we cannot find our apartment key, but know that it has to be
somewhere (because we used it last night to unlock the door), constructively we
can only defend the doubly negated statement
\[ \neg\neg (\exists x\_ \text{the key is at position~$x$}). \]
\end{ex}

@@

\begin{rem}Constructive mathematics is emphatically not the same as computable
mathematics; in particular, there are statements which do admit a witness by a
computer program but which are rejected in (most schools of) constructive
mathematics (see Exercise~\ref{ex:markov}). The apparent tension with the motto
presented above is resolved in its more formal treatment. In any case, the two
subjects are closely related and their connection provides useful
intuition.\end{rem}

@@

By relinquishing the blanket principle of excluded middle, constructive
mathematics allows us to make finer distinctions -- distinctions which are
hidden in classical mathematics, but carry algorithmic and geometric
significance. We have already seen one instance of this increased
expressiveness in Example~\ref{ex:key}: In classical mathematics, we are
trained to cancel double negations as soon as they arise, but constructively
there is a substantial difference between actually having a mathematical object
and merely knowing that its nonexistence is impossible.

\subsection{Decidability of equality}

A further examples pertains to the basic notion of equality. In classical
mathematics, any given elements~$a,b$ of a set~$X$ are either equal or not --
trivially so, by the principle of excluded middle. Constructively, the
situation is more nuanced.

Firstly, in a positive direction, we do have the following result.

\begin{prop}\label{prop:discreteness}
For every pair of natural numbers~$a$ and~$b$, either~$a = b$ or~$a \neq b$.\end{prop}

\begin{proof}[Proof (constructive)]
We spell out only the case~$b = 0$, that is, we only spell out a proof that
every natural number~$a$ is either zero or not. To this end, we do a induction
on~$a$.

The base step~$a = 0$ is trivial, since if~$a = 0$ then also~$a = 0 \vee a \neq
0$ (we can always weaken statements by addition additional disjuncts).

In the inductive step~$a \to a + 1$, we have~$a + 1 \neq 0$ by one of the basic
axioms of arithmetic (for a complete list, see
Exercise~\ref{ex:stability-axioms}) and hence in particular~$a+1 = 0 \vee
a+1\neq0$ by weakening. (This proof is one of the rare instances of a correct
induction proof where the inductive step does not refer to the induction
hypothesis.)
\end{proof}

As from any constructive proof, a computer program witnessing the asserted fact
can be extracted. It will (roughly) have type $\mathsf{Nat} \times \mathsf{Nat}
\to \mathsf{Bool}$, computing~$\mathsf{true}$ or~$\mathsf{false}$ depending on
whether its input agree or are distinct. Because the presented proof proceeded
by induction, the resulting program will proceed by recursion.

Sets like the set of natural numbers for which equality is decidable have a
special name in constructive mathematics, hinting at a connection with topology:
\begin{defn}A set~$X$ is \emph{discrete} if and only if for every
elements~$a,b \in X$, either~$a = b$ or~$a \neq b$.\end{defn}
In classical mathematics, all sets are discrete. Constructively, discreteness
is a nontrivial property, and although~$\NN$ is discrete and with it also~$\ZZ$
and~$\QQ$, there are many important sets which constructively cannot be shown
to be discrete:

\begin{prop}Already in the case~$X = \{\star\}$ it holds that if the
powerset~$P(X)$ is discrete, then the principle of excluded middle
holds.\end{prop}

\begin{proof}Let~$\varphi$ be a statement and consider the element~$K_\varphi
\defeq \{ x \in X \,|\, \varphi \} \in P(X)$. If~$K_\varphi = X$,
then~$\varphi$; and if~$K_\varphi \neq X$, then~$\neg\varphi$.\end{proof}

\begin{prop}If the set of (any of the usual flavors of the) reals is discrete,
then the principle of omniscience holds (XXX this has not been
introduced).\end{prop}

The failure of the discreteness of the set of real numbers can be explained in
computational terms as follows. In computable mathematics, a real numbers is
represented by a program computing better and better rational approximations.
Given two such programs, we can simulate or execute them to obtain
approximations to a desired precision, hoping to distinguish the two
represented real numbers. However, in case that the two numbers actually agree,
in finite time we will never verify this fact: Computably, we cannot rule out
the possibility that a difference between the two numbers will only be detected
in the as yet unexplored range of rational approximations.\footnote{This
argument can be formalized using Turing machines and realizability theory;
see~\cite[Section~4.2.1]{blechschmidt:filmat} for a review. The analysis
dramatically changes (see~\cite[Section~4.2.2]{blechschmidt:filmat}) if we
instead refer to the infinite-time Turing machines of Hamkins and
Lewis~\cite{hamkins-lewis:ittm} which can carry out an infinite amount of
computation before halting. A philosophically intriguing (and, by necessity,
informal) account referring to machines in the real world as opposed to
idealized Turing machines has been put forward by Andrej
Bauer~\cite{bauer:int-mathematics}.}

Lest an impression that only topologically simple domains can be shown to be
discrete in constructive mathematics emerges, here is a more sophisticated
example of a discrete set.

\begin{prop}The set~$\overline{\QQ}$ of algebraic numbers is
discrete.\end{prop}

\begin{proof}See~\cite[Section~XXX]{mines-richman-ruitenburg:constructive-algebra}.\end{proof}

\subsection{Minima of sets of natural numbers}

As an approximate rule of thumb, in the author's personal experience, every
result of classical mathematics which has received constructive scrutiny turned
out to either
\begin{enumerate}
\item be manifestly and unsurprisingly unconstructive, or
\item admit a constructive reformulation.
\end{enumerate}
Of course, the jury is still out on the many results not yet studied from a
constructive point of view.

The results ``every vector space has a basis'' and ``there exist
nonmeasurable subsets of the reals'' are of the first kind. Such results often
provide great abstract insight to the mathematical landscape -- it is
comforting to know that there cannot be vector spaces so bizarre that they
don't have a basis or that the notion of measurable subsets is not trivial.
Such results also tend to not influence concrete situations too much: How could
an infinite basis for which there is no hope that we ever learn any actual
description of inform our computations? (Better uncover topological structure
and look for a Schauder basis!)

Indeed, there are even metatheorems, reported on in
Lecture~\ref{lect:constructive-content}, to the effect that in sufficiently
concrete and logically simple situations, the axiom of choice and the principle
of excluded middle do not allow to prove results which couldn't be proven
without them. And as further evidence, there is Friedman's \emph{grand
conjecture}~\cite{XXX} -- informal but not yet challenged -- that every result
published in the Annals by a mathematician not self-identifying as a set
theorist or logician can be (reformulated to be) provable even in~\textsc{efa},
a foundational system much weaker than~\textsc{zfc},~\textsc{zf} or even Peano
arithmetic.

An example of a result of type~(b) us the basic observation that every
inhabited set of natural numbers contains a minimal element. This result is not
available in constructive mathematics as stated, but there are two constructive
substitutes.

\begin{defn}A set~$X$ is \emph{inhabited} if and only if there exists an
element of~$X$. (Constructively, this is stronger than~$X$ not being
empty.)\end{defn}

\begin{prop}If every inhabited set of natural numbers contains a minimal
element, then the principle of excluded middle holds.\end{prop}

To salvage the classical result, we can strengthen its assumption or weaken its
conclusion.

\begin{defn}A subset~$U \subseteq X$ is \emph{detachable} if and only if for
every element~$x \in X$, either~$x \in U$ or~$x \not\in U$.\end{defn}

For instance, the subset of~$\NN$ consisting of the prime numbers is detachable,
while the subset of those numbers~$n$ such that the~$n$-th Turing machine
terminates cannot constructively be shown to be detachable.\footnote{A
computable witness to detachability of this subset would be a \emph{halting
oracle}, but a basic fact of computability theory is that there no such
oracles.}

\begin{prop}\label{prop:minima}\begin{enumerate}
\item Every detachable inhabited subset of~$\NN$ contains a minimum.
\item Every inhabited subst of~$\NN$ does~\notnot contain a minimum.
\end{enumerate}
\end{prop}

@@

\begin{exercise}[Markov's principle]\label{ex:markov}
Markov's principle is the statement that
\[ \forall f \? \NN \to \{0,1\}\_ (\neg\neg\exists n \? \NN\_ f(n) = 0)
\Rightarrow (\exists n \? \NN\_ f(n) = 0). \]
It is a simple instance of the classical principle of double-negation
elimination, but not available in (most schools of) constructive mathematics.
\begin{enumerate}
\item Why does Markov's principle imply that programs which do not run forever
actually halt?
\item Explain why Markov's principle is witnessed by a computer program (even
though it does not admit a constructive proof). Which assumption on your
metatheory does your argument require?
\end{enumerate}
\end{exercise}

\begin{exercise}[Minima of sets of natural numbers, part one]\label{ex:minima1}
\begin{enumerate}
\item Give a constructive proof of Proposition~\ref{prop:minima}(b).
\item What does the computational witness extracted from the proof of
Proposition~\ref{prop:minima}(a) look like? Can you devise a different proof
corresponding to a different algorithm?
\end{enumerate}
\end{exercise}

@@

\begin{exercise}[Stability of the axioms]\label{ex:stability-axioms}
Peano arithmetic (\textsc{pa}) is set in the language~$(0,S,+,\cdot)$ and has
the following axioms (where leading universal quantifiers are supressed for
brevity):
\begin{enumerate}
\renewcommand{\theenumi}{\arabic{enumi}}
\item $Sx \neq 0$
\item $Sx = Sy \Rightarrow x = y$
\item $y = 0 \vee (\exists x\_ y = Sx)$
\item $x + 0 = x$
\item $x + Sy = S(x+y)$
\item $x \cdot 0 = 0$
\item $x \cdot Sy = (x \cdot y) + x$
\item $P(0) \wedge (\forall n\_ P(n) \Rightarrow P(Sn)) \Longrightarrow
(\forall n\_ P(n))$ (one axiom for each formula~$P(n)$)
\end{enumerate}
Heyting arithmetic~(\textsc{ha}) has exactly the same axioms, but is based on intuitionistic
logic instead of classical logic.
\begin{enumerate}
\item Show that~\textsc{ha} proves the double-negation translation of each
axiom of~\textsc{pa}.
\item Convince yourself that~\textsc{izf} does not prove the double-negation
translation of the axiom of choice. Hence the double-negation translation
alone is insufficient to extract constructive content from proofs using the
axiom of choice.
\end{enumerate}
\end{exercise}

@@

\begin{exercise}[Minima of sets of natural numbers, part two]
In Proposition~\ref{prop:minima}(b) we showed that every inhabited set of natural
numbers does \notnot contain a minimal element.
\begin{enumerate}
\item Redo Exercise~\ref{ex:minima1}(a) using the motto that in proofs of
negated statements, we are allowed to use finitely many instances of the
principle of excluded middle.
\item Explain how Proposition~\ref{prop:minima}(b) follows
from the double-negation embedding applied to the classical fact that every
inhabited set of natural numbers directly contains a minimal element.

{\noindent\scriptsize\emph{Hint.} It is not enough to just cite the
double-negation embedding. The resulting constructive theorem has to be
massaged a bit to yield Proposition~\ref{prop:minima}(b).\par}
\item Explain in computational terms how your proof of
Proposition~\ref{prop:minima}(b) runs its course.
\end{enumerate}
\end{exercise}

@@

Comment on Zorn's lemma
